A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-02-07 rigid_bodies.tex
intro

\begin{abstract}
In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules. Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions. Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
\end{abstract}

@@

\section{Introduction}
Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature
\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
systematic errors
\begin{itemize}
\item[1] Kamberaj\cite{Kamberaj2005}\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
assumption equipartition of the kinetic energy between the translational and rotational degrees of freedom. We demonstrate that this assumption is not valid and it dramatically affects the drift.
to our knowledge, have not been reported in the
order of the operators
\item[2] our paper review
\item[3] coarse grained: 
Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations. This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further. One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

\end{itemize}

In this paper, we report formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases. Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

%The paper is organized as follows. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

>>>>

\begin{abstract}
%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules. Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions. Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
\end{abstract}

@@

\section{Introduction}

Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena. Under those approximations, certain Statistical Mechanics laws might not been necessarily translated into assumptions in Molecular Simulation. For instance, it has already been shown that the unavoidalbe truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and the consequence is that the temperature becomes an undefinite property. In a previous paper,\citep{Abreu_2017} we arrived same now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble. From that observation, we inspected the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, on which two thermostat chains act separately on the rotational and the translational degrees of freedom and it is not able to reproduced de specified temperature.

The importance of rigid body MD stems from the fact that its employment is not limited to molecules modelled as rigid bodies, like the ubiquitous case of water,  but also it increasingly tool as a coarse-grained strategy for sistems designed as colletions of rigid bodies like molecular motors, nanoparticles, proteins, etc. The method of Kamberaj~\textit{et al.}\cite{Kamberaj2005} was actually used in the simulation of small molecules\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}, as well as in molecular machines,\cite{Akimov2008, Konyukhov2010} and nanoparticles.\cite{Knorowski2012, Patra2013}

Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature. It extended phase strategy on which ficticius variables are introduced in order to control the fluctuations of energy and, so the temperature as well. The dynamic of those variables are non-Hamiltonian in nature, which makes the evaluation more difficult, comparing with Hamiltonian dynamics where backward error analysis is done. In fact, it has been used.... In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with SHAKE and Hybrid Monte Carlo (HMC).\cite{Duane1987}  HMC is an exact method in the sense that the truncation erros are not present. And SHAKE is a iterative method that employ lagrange multiplier to solve the holonomic constraints used in MD simulations.For molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.

In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce the numerical integrators. both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which . %We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

The paper is organized as follows. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
